---
title: "Emotions and Temporal Representations"
author: "Enrico Toffalini"
date: "2025-01-10"
include-in-header: assets/header.html
format: 
  html:
    toc: true     
    toc-depth: 2   
    code-summary: "Show the code"
    message: false
    css: custom.css
editor: visual
---

## General Considerations

For simplicity, we will consider only the **behavioral responses**. The primary aim of this study was to investigate whether the Emotion conveyed by a video (Positive, Neutral, Negative) influenced some psychomotor abilities and time processing skills: 

- **Finger tapping**: ability to 1) achieve a lower inter-tap interval (ITI); 2) consistency in ITI (less variation in ITI across the task).
- **Retrospective time estimation**: Accurately estimating the duration of the video just viewed. 
- **Prospective time estimation**: Accurately estimating a one-minute interval during the video, starting when a red dot appears.

Luckily, the authors appreciate open science and transparency, and they have shared their data 🙂 Along with the dataset, they have provided a detailed document explaining its contents. In this case, the file is available here: <a href="Data/README.txt" download>README.txt</a>.

The raw data is divided into two files: 

- <a href="Data/FingerTapping.csv" download>FingerTapping.csv</a>, which contains the data related to the finger-tapping task.
- <a href="Data/TimePerception.csv" download>TimePerception.csv</a>, which includes the data for the two time perception tasks.

## Data Analysis with R

Beginners in R may find helpful resources, including an open-access book and a short course, here: [*Basics of R for Data Science*](https://enricotoffalini.github.io/Basics-R/){target="_blank"}. To get started quickly, check out the [*"Getting Started"*](https://enricotoffalini.github.io/Basics-R/#getting-started){target="_blank"} section, which provides a straightforward guide for installing R and RStudio. Learning R can be a valuable skill for your future, not just for research: it may also give you an advantage in methodological courses and make conducting an empirical thesis easier.

#### Initial Setup: Empty Workspace and Load Packages

```{r, message=F, warning=F}
# empty workspace
rm(list=ls())
# before loading, make sure that a package is installed, otherwise install it with command 
# > install.packages("<package_name>")
library(lme4)
library(lmerTest)
library(glmmTMB)
library(effects)
library(ggplot2)
```

#### Import Data

```{r, message=F, warning=F}
# I placed .csv files into a folder named "Data"
ft = read.csv("Data/FingerTapping.csv")
tp = read.csv("Data/TimePerception.csv")
```

#### Inspect Datasets

Let's see the first few rows of each dataset for figuring out their structures.

```{r, message=F, warning=F}
head(ft)
head(tp)
```

## Finger Tapping

As recommended, let's remove the first trial for each subject. Then, let's inspect the distribution of ITI.

```{r, message=F, warning=F}
ft = ft[ft$Trial != 1, ]
hist(ft$ITI, breaks=100)
```

OUCH! There are a few outliers! Let's remove ITI observations that are too far away from the average:

```{r, message=F, warning=F}
ft = ft[abs((ft$ITI - mean(ft$ITI))/sd(ft$ITI)) < 3, ]
hist(ft$ITI, breaks=100)
```

### Fit Statistical Model on *Finger Tapping*

```{r, message=F, warning=F, cache=T}
# prepare factorial variables
ft$ID = factor(ft$ID)
ft$Emotion = factor(ft$Emotion, levels=c("Neutral","Negative","Positive"))

# actually fit the model
fit_ft = glmmTMB(
  ITI ~ Emotion + (Emotion|ID),
  dispformula = ~ Emotion,
  data = ft,
  family = Gamma(link="log")
)

# see summary of model coefficients
summary(fit_ft)

# plot predicted effects
plot(allEffects(fit_ft))
```

## Time Perception: Retrospective

```{r, message=F, warning=F}
# inspect distribution of Retrospective estimates
hist(tp$Retro_Estimate)
```

Compute (in)accuracy as a ratio:

$$ RetrospectiveError = \frac{\text{Estimate} - \text{ObjectiveTime}}{\text{ObjectiveTime}} $$

```{r, message=F, warning=F}
# compute both absolute and relative precision of estimate
tp$Retro_abs_error = abs((tp$Retro_Estimate - 5) / 5)
tp$Retro_rel_error = (tp$Retro_Estimate - 5) / 5
```

### Fit Statistical Model on *Retrospective estimates*

#### Absolute error

```{r, message=F, warning=F}
# prepare factorial variables
tp$ID = factor(tp$ID)
tp$Emotion = factor(tp$Emotion, levels=c("Neutral","Negative","Positive"))

# actually fit the model - absolute error
fit_re_abs = lmer(
  Retro_abs_error ~ Emotion + (1|ID),
  data = tp
)

anova(fit_re_abs)

# see summary of model coefficients
summary(fit_re_abs)$coefficients

# plot predicted effects
plot(allEffects(fit_re_abs))
```

#### Relative error

```{r, message=F, warning=F}
# actually fit the model - relative error
fit_re_rel = lmer(
  Retro_rel_error ~ Emotion + (1|ID),
  data = tp
)

anova(fit_re_rel)

# see summary of model coefficients
summary(fit_re_rel)$coefficients

# plot predicted effects
plot(allEffects(fit_re_rel))
```

## Time Perception: Prospective

```{r, message=F, warning=F}
# inspect distribution of Retrospective estimates
hist(tp$Prosp_Estimate)
```

OUCH! There is a negative value! Remove it:

```{r, message=F, warning=F}
tp_pro = tp[tp$Prosp_Estimate > 0, ]
hist(tp_pro$Prosp_Estimate)
```

Compute error as a ratio:

$$ ProspectiveError = \frac{\text{Estimate} - \text{ObjectiveTime}}{\text{ObjectiveTime}} $$

```{r, message=F, warning=F}
# compute both absolute and relative error of estimate
tp_pro$Prosp_abs_error = abs((tp_pro$Prosp_Estimate - 60000) / 60000)
tp_pro$Prosp_rel_error = (tp_pro$Prosp_Estimate - 60000) / 60000
```

### Fit Statistical Model on *Prospective estimates*

#### Absolute error

```{r, message=F, warning=F}
fit_pro_abs = lmer(
  Prosp_abs_error ~ Emotion + (1|ID),
  data = tp_pro
)

anova(fit_pro_abs)

# see summary of model coefficients
summary(fit_pro_abs)$coefficients

# plot predicted effects
plot(allEffects(fit_pro_abs))
```

#### Relative error

```{r, message=F, warning=F}
fit_pro_rel = lmer(
  Prosp_rel_error ~ Emotion + (1|ID),
  data = tp_pro
)

anova(fit_pro_rel)

# see summary of model coefficients
summary(fit_pro_rel)$coefficients

# plot predicted effects
plot(allEffects(fit_pro_rel))
```
