---
title: "Emotions and Temporal Representations"
author: "Enrico Toffalini"
date: "2025-01-10"
include-in-header: assets/header.html
format: 
  html:
    toc: true     
    toc-depth: 2   
    code-summary: "Show the code"
    message: false
    css: "assets/custom.css"
editor: visual
---

## General Considerations

For simplicity, we will consider only the **behavioral responses**. The primary aim of this study was to investigate whether the Emotion conveyed by a video (Positive, Neutral, Negative) influenced some psychomotor abilities and time processing skills:

-   **Finger tapping**: ability to 1) achieve a lower inter-tap interval (ITI); 2) consistency in ITI (less variation in ITI across the task).
-   **Retrospective time estimation**: Accurately estimating the duration of the video just viewed.
-   **Prospective time estimation**: Accurately estimating a one-minute interval during the video, starting when a red dot appears.

Luckily, the authors appreciate open science and transparency, and they have shared their data 🙂 Along with the dataset, they have provided a detailed document explaining its contents. In this case, the file is available here: <a href="Data/README.txt" download>README.txt</a>.

The raw data is divided into two files:

-   <a href="Data/FingerTapping.csv" download>FingerTapping.csv</a>, which contains the data related to the finger-tapping task.
-   <a href="Data/TimePerception.csv" download>TimePerception.csv</a>, which includes the data for the two time perception tasks.

## Data Analysis with R

Beginners in R may find helpful resources, including an open-access book and a short course, here: [*Basics of R for Data Science*](https://enricotoffalini.github.io/Basics-R/){target="_blank"}. To get started quickly, check out the [*"Getting Started"*](https://enricotoffalini.github.io/Basics-R/#getting-started){target="_blank"} section, which provides a straightforward guide for installing R and RStudio. Learning R can be a valuable skill for your future, not just for research: it may also give you an advantage in methodological courses and make conducting an empirical thesis easier.

#### Initial Setup: Empty Workspace and Load Packages

```{r, message=F, warning=F}
# empty workspace
rm(list=ls())
# before loading, make sure that a package is installed, otherwise install it with command 
# > install.packages("<package_name>")
library(lme4)
library(lmerTest)
library(glmmTMB)
library(effects)
library(ggplot2)
```

#### Import Data

```{r, message=F, warning=F}
# I placed .csv files into a folder named "Data"
ft = read.csv("Data/FingerTapping.csv")
tp = read.csv("Data/TimePerception.csv")
```

#### Inspect Datasets

Let's see the first few rows of each dataset for figuring out their structures.

```{r, message=F, warning=F}
head(ft)
head(tp)
```

## Finger Tapping

As recommended, let's remove the first trial for each subject. Then, let's inspect the distribution of ITI.

```{r, message=F, warning=F}
ft = ft[ft$Trial != 1, ]
hist(ft$ITI, breaks=100)
```

OUCH! There are a few outliers! Let's remove ITI observations that are too far away from the average:

```{r, message=F, warning=F}
ft = ft[abs((ft$ITI - mean(ft$ITI))/sd(ft$ITI)) < 3, ]
hist(ft$ITI, breaks=100)
```

#### Fit Statistical Model on *Finger Tapping*

Because each participant provides multiple responses, we can test not only if Emotion affects average ITI, but also if it affects its dispersion. Let's run a special model for this:

```{r, message=F, warning=F, cache=T}
# prepare factorial variables
ft$ID = factor(ft$ID)
ft$Emotion = factor(ft$Emotion, levels=c("Neutral","Negative","Positive"))

# actually fit the model
fit_ft = glmmTMB(
  ITI ~ Emotion + (Emotion|ID),
  dispformula = ~ Emotion,
  data = ft,
  family = Gamma(link="log")
)

# see summary of model coefficients
summary(fit_ft)

# plot predicted effects
plot(allEffects(fit_ft))
```

## Time Perception: Retrospective

```{r, message=F, warning=F}
# inspect distribution of Retrospective estimates
hist(tp$Retro_Estimate)
```

Compute (in)accuracy as a ratio:

$$ RetrospectiveError = \frac{\text{Estimate} - \text{ObjectiveTime}}{\text{ObjectiveTime}} $$

```{r, message=F, warning=F}
# compute precision (error) of retrospective estimate
tp$Retro_Error = (tp$Retro_Estimate - 5) / 5
```

In fact, this does not reflect the amount of "error" in the estimation: positive values indicate *overestimation*, while negative values indicate *underestimation*. If a particular video is largely overestimated by some and largely underestimated by others, its mean estimate will be "zero", but this does NOT indicate accuracy. However, accuracy vs inaccuracy might be reflected by a smaller vs larger dispersion of estimate scores.

#### Fit Statistical Model on *Retrospective estimates*

```{r, message=F, warning=F}
# prepare factorial variables
tp$ID = factor(tp$ID)
tp$Emotion = factor(tp$Emotion, levels=c("Neutral","Negative","Positive"))

# actually fit the model
fit_re = glmmTMB(
  Retro_Error ~ Emotion + (1|ID),
  dispformula = ~ Emotion,
  data = tp
)

# see summary of model coefficients
summary(fit_re)

# plot predicted effects
plot(allEffects(fit_re))
```

## Time Perception: Prospective

```{r, message=F, warning=F}
# inspect distribution of Retrospective estimates
hist(tp$Prosp_Estimate)
```

OUCH! There is a negative value! Remove it:

```{r, message=F, warning=F}
tp$Prosp_Estimate[tp$Prosp_Estimate < 0] = NA
hist(tp$Prosp_Estimate)
```

Compute error as a ratio:

$$ ProspectiveError = \frac{\text{Estimate} - \text{ObjectiveTime}}{\text{ObjectiveTime}} $$

```{r, message=F, warning=F}
# compute precision (error) of prospective estimate
tp$Prosp_Error = (tp$Prosp_Estimate - 60000) / 60000
```

#### Fit Statistical Model on *Prospective estimates*

```{r, message=F, warning=F}

fit_pro = glmmTMB(
  Prosp_Error ~ Emotion + (1|ID),
  dispformula = ~ Emotion,
  data = tp
)

# see summary of model coefficients
summary(fit_pro)

# plot predicted effects
plot(allEffects(fit_pro))
```

## Comments and Conclusions:

-   For **Finger tapping**, it seems that *Negative* Emotion increases both average ITI and its dispersion within participants.
-   For the **Retrospective task**, it seems that Emotion has not particular effect on either estimation or its dispersion.
-   For the **Prospective task**, it seems that Emotion affects accuracy, causing both an average overestimation of duration in the *Negative* condition and some increase in its dispersion (absolute error).

[An important point, when computing statistical analyses and finding relevant/significant effects, is to estimate how large they are. Since we used variables on interpretable metrics (ratio of error, milliseconds of ITI) the interpretation might be straightforward. However, standardized effect sizes (e.g., [*Cohen's d*](https://en.wikiversity.org/wiki/Cohen%27s_d){target="_blank"}) are also recommended.]{style="font-size:14px"}
