---
title: "Physiological Data Analysis"
author: "Enrico Toffalini"
date: "2025-02-11"
include-in-header: assets/header.html
format: 
  html:
    toc: true     
    toc-depth: 3   
    code-summary: "Show the code"
    message: false
    css: "assets/custom.css"
editor: visual
---

```{r, message=F, warning=F, echo=F}
rm(list=ls())
library(glmmTMB)
library(effects)
phy = read.csv("Data/PhysiologicalDataProcessed.csv")
phy_raw_sample = read.csv("Data/PhysiologicalRawData_sample.csv")

phy$Condition = factor(phy$Condition, levels=c("Resting","Neutral","Positive","Negative"))
```

## Raw Data

#### Dataset inspect

```{r, message=F, warning=F}
head(phy_raw_sample)
```

#### Plot ECG sample (first 5 seconds)

*Exercise:* Based on the plot below, could you provide an estimate of *beats per minute*?

```{r, message=F, warning=F}
#| code-fold: true
library(ggplot2)

ggplot(phy_raw_sample[1:2500,],aes(x=second,y=ECG))+
  geom_line()+
  scale_x_continuous(breaks=seq(0,100,.5))+
  theme(text=element_text(size=18))+
  ylab("ECG (microvolt)")
```

#### Plot EDA sample (first 5 seconds), with a *gam* smoother

```{r, message=F, warning=F}
#| code-fold: true
ggplot(phy_raw_sample[1:2500,],aes(x=second,y=EDA))+
  geom_line()+
  scale_x_continuous(breaks=seq(0,100,.5))+
  geom_smooth(linewidth=3)+
  theme(text=element_text(size=18))+
  ylab("ECG (microvolt)")
```

## Processed Data

#### Dataset inspect

```{r, message=F, warning=F}
phy[1:10,]
```

### Heart Rate (HR)

#### HR visualization - by subject

```{r, message=F, warning=F}
ggplot(phy,aes(x=Minute,y=HR,color=Condition))+
  geom_line(aes(x=Minute,y=HR,group=paste0(ID,Condition)),alpha=.6,linewidth=1)+
  theme(text=element_text(size=18))+
  ylab("HR (bpm)")
```

#### HR visualization - aggregated and smoothed

```{r, message=F, warning=F}
ggplot(phy,aes(x=Minute,y=HR,color=Condition,fill=Condition))+
  geom_smooth(alpha=.2,linewidth=2)+
  theme(text=element_text(size=18))+
  ylab("HR (bpm)")
```

#### Statistical models on HR

Due to serious convergence issues, I cannot enter a full random effects formula, so results must be taken with caution.

```{r, message=F, warning=F, cache=T}
fit_hr_Multiplicative = glmmTMB(
  HR ~ Minute * Condition + (1|ID),
  dispformula = ~ Condition,
  data = phy
)
fit_hr_Additive = glmmTMB(
  HR ~ Minute + Condition + (1|ID),
  dispformula = ~ Condition,
  data = phy
) 

anova(fit_hr_Multiplicative, fit_hr_Additive)
```

```{r, message=F, warning=F}
# see summary of model coefficients
summary(fit_hr_Additive)

# plot predicted effects
plot(allEffects(fit_hr_Additive))
```

### Skin Conductance Response (SCR)

#### SCR visualization - by subject

```{r, message=F, warning=F}
ggplot(phy,aes(x=Minute,y=SCR,color=Condition))+
  geom_line(aes(x=Minute,y=SCR,group=paste0(ID,Condition)),alpha=.6,linewidth=1)+
  theme(text=element_text(size=18))+
  ylab("SCR (microsiemens)")
```

There are extreme outlier observations! Let's remove them:

```{r, message=F, warning=F}
phy$SCR[phy$SCR>600] = NA
ggplot(phy,aes(x=Minute,y=SCR,color=Condition))+
  geom_line(aes(x=Minute,y=SCR,group=paste0(ID,Condition)),alpha=.6,linewidth=1)+
  theme(text=element_text(size=18))+
  ylab("SCR (microsiemens)")
```

#### SCR visualization - aggregated and smoothed

```{r, message=F, warning=F}
ggplot(phy,aes(x=Minute,y=SCR,color=Condition,fill=Condition))+
  geom_smooth(alpha=.2,linewidth=2)+
  theme(text=element_text(size=18))+
  ylab("SCR (microsiemens)")
```

#### Statistical models on SCR

```{r, message=F, warning=F, cache=T}
fit_scr_Multiplicative = glmmTMB(
  SCR ~ Minute * Condition + (1|ID),
  dispformula = ~ Condition,
  family = gaussian(link="log"),
  data = phy
)
fit_scr_Additive = glmmTMB(
  SCR ~ Minute + Condition + (1|ID),
  dispformula = ~ Condition,
  family = gaussian(link="log"),
  data = phy
) 

anova(fit_scr_Multiplicative, fit_scr_Additive)
```

```{r, message=F, warning=F}
# see summary of model coefficients
summary(fit_scr_Additive)

# plot predicted effects
plot(allEffects(fit_scr_Additive))
```

## Comments and Conclusions

-   For **Heart Rate**, we observed highest mean values (and highest dispersion of scores) during the *Resting* state, while there were no relevant changes across the *Neutral*, *Positive*, or *Negative* video conditions, somehow unexpectedly. There was a slight tendency for HR to increase across time, independently from Condition.

-   For **Skin Conductance Response**, we observed highest mean values during the *Positive* video condition, followed by *Negative* and *Neutral* conditions, with *Resting* state presenting the lowest values. Dispersion of scores also followed this patter. SCR also tended to reduce across time within each Condition, and quite independently from Condition.
